---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
--- 


## Loading and preprocessing the data
#### Load libraries
```{r, results='hide', message=FALSE}
library(dplyr)
library(ggplot2)
```
#### Load data
```{r}
steps.df <- read.csv("activity.csv")
```
#### Structure of data
```{r}
str(steps.df)
head(steps.df)
```
#### Convert the date column from factor to POSIXlt type
```{r}
steps.df$date <- as.POSIXct(strptime(steps.df$date, format="%Y-%m-%d"))
```
## What is mean total number of steps taken per day?
#### Total steps taken, grouped by date
```{r, results='hide'}
steps.bydate <- summarise(group_by(steps.df, date), total = sum(steps, na.rm=T))
```
#### Histogram of total steps taken per day
```{r}
qplot(total, data=steps.bydate, binwidth=500) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Count") + 
  ggtitle("Histogram of Steps Taken")
```

#### Mean and median of total steps taken per day
Mean is `r format(round(mean(steps.bydate$total), 2), scientific=4)` and Median is `r format(round(median(steps.bydate$total), 2), scientific=4)`. 

## What is the average daily activity pattern?
#### Average steps taken by Interval across all days
```{r, results='hide'}
steps.byinterval <- summarise(group_by(steps.df, interval), Mean = mean(steps, na.rm=T))
```
#### Time series plot of Average Steps taken by Interval
```{r}
qplot(interval, Mean, data=steps.byinterval, geom="line") + 
  xlab("Interval") + 
  ylab("Average Steps Taken") + 
  ggtitle("Time Series of Average Steps Taken")
```

#### Time series plot of Log(Average Steps taken) by Interval
```{r}
qplot(interval, log(Mean + 1), data=steps.byinterval, geom="line") + 
  xlab("Interval") + 
  ylab("Log(Average Steps Taken)") + 
  ggtitle("Time Series of Log(Average Steps Taken)")
```

#### Interval containing maximum number of average steps
The interval containing the maximum number of average steps, ```r max(steps.byinterval$Mean)```,  is ```r steps.byinterval[steps.byinterval$Mean == max(steps.byinterval$Mean), ]$interval```.

## Imputing missing values
#### Missing values in the data
* Total number of rows in the data is `r nrow(steps.df)`. 
* Number of rows with missing Steps is `r sum(is.na(steps.df$steps))`. 
* Number of rows with missing Interval is `r sum(is.na(steps.df$interval))`. 
* Number of rows with missing Date is `r sum(is.na(steps.df$date))`.

#### Impute missing Steps values by replacing it with the average number of steps for that specific interval
```{r}
steps.imputed <- mapply(function(st, itv)  { ifelse(is.na(st), 
                                                      subset(steps.byinterval, 
                                                             interval==itv, 
                                                             select=c("Mean"))$Mean, 
                                                      st)},
                           steps.df$steps, 
                           steps.df$interval)

```
#### Create a new data set with the missing steps filled in
```{r}
steps.df.imputed <- data.frame(steps = steps.imputed, interval=steps.df$interval, date=steps.df$date)
```
#### Total steps taken, grouped by date
```{r, results='hide'}
steps.imputed.bydate <- summarise(group_by(steps.df.imputed, date), total = sum(steps, na.rm=T))
```
#### Histogram of total steps taken per day
```{r}
qplot(total, data=steps.imputed.bydate, binwidth=500) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Count") + 
  ggtitle("Histogram of Steps Taken (after imputation)")
```

#### Mean and median of total steps taken per day
Mean is `r format(round(mean(steps.imputed.bydate$total), 2), scientific=4)` and Median is `r format(round(median(steps.imputed.bydate$total), 2), scientific=4)`. 
After imputation, the mean increased by `r format(round(mean(steps.imputed.bydate$total) - mean(steps.bydate$total), 2), scientific=4)` steps per day and the median increased by `r format(round(median(steps.imputed.bydate$total) - median(steps.bydate$total), 2), scientific=4)`.
Since the median and mean are equal, the new distribution of steps by date seems to be symmetric around the mean.

## Are there differences in activity patterns between weekdays and weekends?
#### Create a new factor variable 'daytype'
```{r, results='hide'}
steps.df.imputed$daytype = as.factor(ifelse(strftime(steps.df.imputed$date, format="%a") %in% c("Sat", "Sun"), 'weekend', 'weekday'))
```

#### Find the average steps taken per interval 
```{r, results='hide'}
steps.imputed.byinterval <- summarise(group_by(steps.df.imputed, daytype, interval), Mean = mean(steps))
```

#### Time series plot for average number of steps taken for weekend and weekdays
```{r}
qplot(interval, Mean, data=steps.imputed.byinterval, geom="line") +
     facet_wrap( ~daytype, ncol=1) +
     xlab("Interval") + 
     ylab("Average Steps Taken") + 
     ggtitle("Time Series of Average Steps Taken By Day Type")
```
